Normalization by Evaluation for \\ Intuitionistic Propositional Logic

### 2018-07-18 notes/nbeSum.tex
ITC: reify/reflect for disjunction and absurdity

\para{Formulas} and hypotheses lists (contexts).
\[
\begin{array}{lllrll}
  \Atom & \ni & P,Q & & & \mbox{atomic propositions} \\
  \Form & \ni & A,B,C,\dots
     & ::= & \top \mid A \land B \mid A \imp B
       & \mbox{negative formulas} \\
     &&& \mid & \bot \mid A \lor B \mid P
       & \mbox{positive formulas} \\
\\[1ex]
  \Cxt & \ni & \Gamma,\Delta & ::= & \cempty & \mbox{empty context}
\\ &&& \mid & \Gamma.A & \mbox{context extension}
\end{array}
\]
Indices $x$ into the context $\Gamma$ of hypothesis, locating hypothesis
$A$, are written $x : \Hyp[\Gamma]A$ and defined inductively by the
following constructors:
\[
  \ru{}{\ttop : \Hyp[\Gamma.A]A}
\qquad
  \ru{x : \Hyp[\Gamma]A
    }{\pop[B] x : \Hyp[\Gamma.B]A}
\]
Deriviations $t$ of the truth of a formula $A$ under assumptions
$\Gamma$ (judgement $\Gamma \der A$)
are written $t : \Tm[\Gamma]A$ and defined inductively as follows:

@@

The interpretation $\den{A \imp B}$ of implication will follow the
Brouwer-Heyting-Kolmogorov (BHK) interpretation of intuitionistic
logic:
\begin{center}\itshape
  A proof of $A \imp B$ is a method turning any proof of $A$ into a
  proof of $B$.
\end{center}
However, the direct lifting of the functions space $\den{A \imp
  B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
not give a presheaf because of the contravariant occurrence of
presheaf $\den A$.  Instead, we force monotonicity by quantifying over
all OPEs of $\Gamma$:
\[
  \den{A \imp B}_\Gamma = \prod_{\Delta : \Cxt} \prod_{\tau : \Delta \leq \Gamma} (\den A
  _\Delta \to \den B _\Delta)
\]
Reification is also challenging.  Let $f : \den{A \imp B}_\Gamma$ and
attempt
%$\tau : \Delta \leq \Gamma$
\[
  \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
    = \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
\]
where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
which intuitively should represent the new hypothesis $\top :
\Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural
transformation $\Hyp A \todot \den A$ which \emph{reflects} variables
into the semantics.  In order to define it by
induction on $A$, we need to generalize it to
\[
  \treflect^A : \Ne A \todot \den A
  .
\]
Let us make sure we can define it at atoms and conjunctions:
\[
\begin{array}{lllll}
  \treflect^P & = & \tne & : & \Ne P \todot \Nf P \\
  \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
  \treflect^{A \land B} & = & [\, \treflect^A \comp \andE1  , \;
                             \treflect^B \comp \andE2 ]
    & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
\end{array}
\]
Observe the eliminations $\andE i$ introduced to reflect conjunctive hypotheses!

@@

\subsection{Digression: The CCC of reifyable presheaves}

@@

\section{Extension to the positive connectives}

>>>>

\para{Formulas} and hypotheses lists (contexts).
\[
\begin{array}{lllrll}
  \Atom & \ni & P,Q & & & \mbox{atomic propositions} \\
  \Form & \ni & A,B,C,\dots
     & ::= & \top \mid A \land B \mid A \imp B
       & \mbox{negative formulas} \\
     &&& \mid & \bot \mid A \lor B \mid P
       & \mbox{positive formulas} \\
\\[1ex]
  \Cxt & \ni & \Gamma,\Delta & ::= & \cempty & \mbox{empty context}
\\ &&& \mid & \Gamma.A & \mbox{context extension}
\end{array}
\]
We write just $A$ for the singleton context $\cempty.A$.

Indices $x$ into the context $\Gamma$ of hypothesis, locating hypothesis
$A$, are written $x : \Hyp[\Gamma]A$ and defined inductively by the
following constructors:
\[
  \ru{}{\ttop : \Hyp[\Gamma.A]A}
\qquad
  \ru{x : \Hyp[\Gamma]A
    }{\pop[B] x : \Hyp[\Gamma.B]A}
\]
Deriviations $t$ of the truth of a formula $A$ under assumptions
$\Gamma$ (judgement $\Gamma \der A$)
are written $t : \Tm[\Gamma]A$ and defined inductively as follows:

@@

The interpretation $\den{A \imp B}$ of implication will follow the
Brouwer-Heyting-Kolmogorov (BHK) interpretation of intuitionistic
logic:
\begin{center}\itshape
  A proof of $A \imp B$ is a method turning any proof of $A$ into a
  proof of $B$.
\end{center}
However, the direct lifting of the functions space $\den{A \imp
  B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
not give a presheaf because of the contravariant occurrence of
presheaf $\den A$.  Instead, we force monotonicity by quantifying over
all OPEs of $\Gamma$:
\[
  \den{A \imp B}_\Gamma = \prod_{\Delta : \Cxt} \prod_{\tau : \Delta \leq \Gamma} (\den A
  _\Delta \to \den B _\Delta)
\]
Reification is also challenging.  Let $f : \den{A \imp B}_\Gamma$ and
attempt
%$\tau : \Delta \leq \Gamma$
\[
  \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
    = \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
\]
where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
which intuitively should represent the new hypothesis $\top :
\Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural
transformation $\Hyp A \todot \den A$ which \emph{reflects} variables
into the semantics.  In order to define it by
induction on $A$, we need to generalize it to
\[
  \treflect^A : \Ne A \todot \den A
  .
\]
Let us make sure we can define it at atoms and conjunctions:
\[
\begin{array}{lllll}
  \treflect^P & = & \tne & : & \Ne P \todot \Nf P \\
  \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
  \treflect^{A \land B} & = & \langle\, \treflect^A \comp \andE1  , \;
                             \treflect^B \comp \andE2 \rangle
    & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
\end{array}
\]
Observe the eliminations $\andE i$ introduced to reflect conjunctive hypotheses!

@@

\subsection{Digression: The CCC of reifiable presheaves}

@@

%\section{Extension to the positive connectives}
\section{Disjunction and absurdity}

For sets $S_1,S_2 : \SET$ let $S_1 + S_2$ denote the disjoint union with
injections $\inj{i} : S_i \to S_1 + S_2$ and case distinction
distinction $[f_1,f_2] : S_1 + S_2 \to T$ for $f_i : S_i \to T$
such that $[f_1,f_2] \comp \inj i = f_i$.

Presheaf coproduct is given pointwise by $(\A \hatplus \B)_\Gamma = \A_\Gamma + \B_\Gamma$.
\begin{exercise}
  Show that $\hatplus$ is indeed a coproduct in $\PSh$.
\end{exercise}

The first idea how to model disjunction,
$\den{A \lor B}_\Gamma = \den{A}_\Gamma + \den{B}_\Gamma$, fails
because we cannot reflect disjunctive hypotheses.  \Eg, an element of
$\den{A \lor B}_{A \lor B}$ should be of the form $\inj1\,a$ or
$\inj2\,b$, but the decision on first or second injection cannot be
made here;  it depends on the hypothesis $A \lor B$.  The solution is
to extend the semantics to allow case distinction on variables (and
neutrals) before making the decisions needed for disjunctions.

Given a presheaf $\A$, we define a new presheaf $\Cover \A$
inductively by the following constructors:
\begin{gather*}
  \ru{a : \A_\Gamma}{\returnC a : \Cover\A_\Gamma}
\qquad
  \ru{t : \NeG \bot}{\falseC t : \Cover\A_\Gamma}
\\[1ex]
  \ru{t : \NeG{C \lor D} \qquad
      c : \Cover\A_{\Gamma.C} \qquad
      d : \Cover\A_{\Gamma.D}
    }{\orC t c d : \Cover\A_\Gamma}
\end{gather*}
\begin{exercise}
  Show that $\Cover\A$ is indeed a presheaf.
\end{exercise}
We can think of $c : \Cover\A_\Gamma$ as a decision tree with leaves in $\A$
and nodes labelled by terms we case on.  Nodes of type $\tfalseC$ have
no subtrees since \emph{ex falsum quod libet}, and $\torC$ nodes
casing on $t : \NeG{C \lor D}$ have
two subtrees, one having a new hypothesis $C$, and one having a new
hypothesis $C$.

We can replace all the leaves according to a function $f : \A \todot \B$,
this makes $\tCover$ a \emph{functor} in the category of presheaves.  We can
also replace all the leaves by new case trees via a natural transformation
\[
   \tjoinC ~:~ \Cover{\Cover\A} \todot \Cover\A
\]
which makes $\tCover$ a \emph{monad} in $\PSh$.
\begin{exercise}
  Define the functorial action $\Cover f$ and $\tjoinC$ and prove the
  monad laws.
\end{exercise}
Further, if the leaves of a case tree are normal forms, we can turn it into one big
normal form via:
\[
  \tpasteNf ~:~ \Cover{\Nf A} \todot \Nf A
\]
\begin{exercise}
  Define $\tpasteNf$.
\end{exercise}

\noindent
We have everything now to model disjunction:
\[
\begin{array}{lcl}
\den{A \lor B} & = & \Cover{\den A \hatplus \den B}
\\[2ex]
\treflect^{A \lor B} & : & \Ne{A \lor B} \todot \den{A \lor B} \\
\reflect[\Gamma]{A \lor B}t & = & \orC t
  {(\returnCP {\injP1 {\reflect[\Gamma.A]A{(\hyp \ttop)}}})}
  {\\ && \qquad\ (\returnCP {\injP2 {\reflect[\Gamma.B]B{(\hyp \ttop)}}})}
\\[2ex]
\treifyOr^{A \lor B} & : & \den{A} \hatplus \den{B} \todot \Nf{A \lor B} \\
\reifyOr[\Gamma]{A \lor B}(\inj1\, a) & = & \orIP1 {\reify[\Gamma]A a} \\
\reifyOr[\Gamma]{A \lor B}(\inj2\, b) & = & \orIP2 {\reify[\Gamma]B b}
\\[1.5ex]
\treify^{A \lor B} & : & \den{A \lor B} \todot \Nf{A \lor B} \\
\treify^{A \lor B} & = & \tpasteNf \comp \Cover{\treifyOr^{A \lor B}} \\
\end{array}
\]
Reification works by first reifying the leaves of the case tree using
$\treifyOr$, and then turning the whole tree into a normal form using $\tpasteNf$.
\begin{exercise}
  Compute $\reify[P \lor Q]{P \lor Q} (\reflect[P \lor Q]{P \lor Q}
  (\hyp \ttop))$.
\end{exercise}
For absurdity, we use the presheaf $\hat0_\Gamma = \{\}$ with initial
morphism $\botelim : \hat0 \todot \C$.
\[
\begin{array}{lcl}
\den{\bot} & = & \Cover{\hat0}
\\[2ex]
\treflect^{\bot} & : & \Ne{\bot} \todot \den{\bot} \\
\reflect[\Gamma]{\bot}t & = & \falseC t
\\[2ex]
\treify^{\bot} & : & \den{\bot} \todot \Nf{\bot} \\
\treify^{\bot} & = & \tpasteNf \comp \Cover{\botelim} \\
\end{array}
\]
To reify a value $c : \den\bot$, observe that because $\hat0$ is
empty, $c$ can only be a case tree without leaves, \ie, all branches
end in $\tfalseC$ nodes.  Thus, the mapping
$\Cover{\botelim} : \Cover{\hat0} \todot \Cover{\Nf\bot}$ is merely a
type cast.  The subsequent $\tpasteNf$ will turn the leafless tree $c$
into a normal form consisting entirely of splits $\torE$ and $\tfalseE$.

\begin{exercise}
Prove that $\treify^{A} \comp \treflect^{A}$ produces the long
$\eta$-normal form if applied to variables (\ie, hypotheses).
\end{exercise}

Evaluation poses one more challenge.
